Categorical MSM

### 2017-06-20 categorical_msm.Rmd
updated

```{r libraries, message=F}
library(tidyverse)
library(broom)
library(contrast)
```

@@

# create normal distribution of x with mean of 16 and sd of 4
sim_data <- data_frame(x = rnorm(n, 16, 4)) %>% 
  # simulate y data with linear logit relationship to x
  mutate(y = rbinom(n, 1, 1/(1+exp(-(-1+0.10*x)))),
         # split in to three categories
         x_cat = ifelse(x < quantile(x,1/3), 1,
                 ifelse(x > quantile(x,2/3), 3, 2)),
         # dummy variable
         x1 = ifelse(x_cat == 1, 1, 0),
         x2 = ifelse(x_cat == 2, 1, 0),
         x3 = ifelse(x_cat == 3, 1, 0),
         x2_3 = ifelse(x_cat == 2 | x_cat == 3, 1, 0))
```

@@

# estimates good enough for example
mod1 <- glm(y~x, sim_data, family="binomial"(link="logit"))
mod1$coefficients

# plot predicted probability of the linear funciton
pr_y_x <- predict(mod1, sim_data, type="response")

@@

Standard analysis asking what the risk difference is strata to strata

```{r risk difference}
# 2x3 table
y_x <- xtabs(~x_cat+y, sim_data)
pr_yx <- prop.table(y_x, 1)
pr_yx
# difference in risk from strata 2 to 1
pr_yx[2,2] - pr_yx[1,2]
# difference in risk from strata 3 to 1
pr_yx[3,2] - pr_yx[1,2]
# difference in risk from strata 3 to 2
pr_yx[3,2] - pr_yx[2,2]

# estimate risk difference using binomial, link=identity for appropriate SE
risk_diff_mod <- glm(y~x2+x3, sim_data, family="binomial"(link="identity"))
summary(risk_diff_mod)

# comparing difference in x3 to x2
risk_diff_mod2 <- glm(y~x3+x1, sim_data, family="binomial"(link="identity"))
summary(risk_diff_mod2)
# difference between groups 3 and 2 as well
# now what if we combine groups x 3 and 2 together and compared against 1

risk_diff_mod3 <- glm(y~x2_3, sim_data,family="binomial"(link="identity"))
summary(risk_diff_mod3)

>>>>

```{r libraries, message=F}
library(tidyverse) 
library(broom) # for tidy model output and bootstrapping
library(MSModelR) # my package for simple msm functions
library(contrast)
```

@@

# create normal distribution of x with mean of 16 and sd of 4
sim_data <- data_frame(x = rnorm(n, 16, 4)) %>% 
  # simulate y data with linear logit relationship to x
  mutate(y = rbinom(n, 1, 1/(1+exp(-(-1+0.10*x)))),
         # split in to three categories
         x_cat = ifelse(x < quantile(x,1/3), 0,
                 ifelse(x > quantile(x,2/3), 2, 1)),
         # dummy variable
         x0 = ifelse(x_cat == 0, 1, 0),
         x1 = ifelse(x_cat == 1, 1, 0),
         x2 = ifelse(x_cat == 2, 1, 0),
         x1_2 = ifelse(x_cat == 1 | x_cat == 2, 1, 0))
```

Truth.

@@

# estimates good enough for example
true_mod <- glm(y~x, sim_data, family="binomial"(link="logit"))
tidy(true_mod)

# plot predicted probability of the linear funciton
pr_y_x <- predict(true_mod, sim_data, type="response")

@@

Standard analysis asking what the risk ratio is strata to strata

```{r risk ratio}
# 2x3 table
y_x <- xtabs(~x_cat+y, sim_data)
pr_yx <- prop.table(y_x, 1)
pr_yx
# difference in risk from strata 2 to 1
pr_yx[2,2] - pr_yx[1,2]
# difference in risk from strata 3 to 1
pr_yx[3,2] - pr_yx[1,2]
# difference in risk from strata 3 to 2
pr_yx[3,2] - pr_yx[2,2]

# estimate risk ratio
risk_ratio_mod <- glm(y~x1+x2, sim_data, family="binomial"(link="logit"))
tidy(risk_ratio_mod)

# comparing difference in x2 to x1
risk_ratio_mod2 <- glm(y~x2+x0, sim_data, family="binomial"(link="logit"))
tidy(risk_ratio_mod2)

# treating the categorical variables as continous
risk_ratio_mod3 <- glm(y~x_cat, sim_data, family="binomial"(link="logit"))
tidy(risk_ratio_mod3)

# now what if we combine groups x 3 and 2 together and compared against 1
risk_ratio_mod4 <- glm(y~x1_2, sim_data,family="binomial"(link="logit"))
tidy(risk_ratio_mod4)
```

Loss of information by combining the mid and high strata. But would this truly reflect a population effect where we may ask the question what if we reduced to background levels represented by strata 0? Something to think about.

Trying out MSM estimates.
